---
title: "2025_Dahl&Edmunds_H1 code"
output: html_document
date: "2025-06-28"
---

#load libraries
```{r, libraries, echo = FALSE, message = FALSE, warning = FALSE}
# Load all required libraries
library(tidyverse)        # includes dplyr, ggplot2, readr, tibble, stringr, etc.
library(vegan)            # for nMDS (metaMDS), envfit, diversity analysis
library(phyloseq)         # estimate_richness, otu_table (Chao1 index)
library(reshape2)         # melt (long-format transformations)
library(ggpmisc)          # stat_poly_line(), stat_poly_eq()
library(sjPlot)           # plot_model() and effects visualization
library(sjmisc)           # supporting functions for sjPlot
library(sjlabelled)       # labeling support for sjPlot
library(gt)               # create styled summary tables
library(grid)             # low-level grob rendering (used with gt table images)
library(gridExtra)        # arranging plots/grobs (optional but useful)
library(cowplot)          # combining plots (e.g., draw_plot)
library(viridis)          # color scales (e.g., in nMDS plots)
library(ggrepel)          # non-overlapping text labels on plots
library(magick)           # reading/processing images (e.g., gt table as PNG)
library(gtsummary)        # optional: For cleaner summary tables
library(car)              # outlierTest()

# Optional: Set a clean default ggplot theme
theme_set(theme_classic(base_size = 15))
```

# H1 - Patch size and proximity to neighbors affects benthic richness 
# Size 
## read in and clean data
### Moorea
```{r, Moorea bommie data}
## need to extract Bray Curtis matrices which will be placed in the nearest neighbor dataset (read in below)
MooreaH1 <- read_csv(here("Data","Moorea_H1.csv"), 
                          show_col_types = FALSE) %>%
  mutate(site=as.factor(site))

#log matrix for diversity/richness indices
MproxmatrixLOG <- log(MooreaH1[,-1:-13]+1)
mooreamatrix <- MooreaH1[,-1:-13]


#switch to vertical orientation
MooreaH1VERT <- melt(MooreaH1, id=c("ID", "ID2", "site", "l", "w", "h", "c", "SAcm", "SAm", "Rl", "Rw", "Clength", "Cwidth" ), variable.name = "species")

MooreaH1VERT <- MooreaH1VERT %>% 
  mutate(value = ifelse(is.na(value), 0, value))

```

### St. John
```{r, St. John bommie data}

StJohnH1 <- read_csv(here("Data","StJohn_H1.csv"), 
                          show_col_types = FALSE) %>%
  mutate(site=as.factor(site))

#log matrix for diversity.richness indices
STJproxmatrixLOG <- log(StJohnH1[,-1:-13]+1)
stJmatrix <- StJohnH1[,-1:-13]

```

### Chao1 richness calculation
#### Moorea
```{r, chao 1, echo = FALSE, message=FALSE, warning=FALSE}
#chao1 richness on log + 1 transformed data
matrix(mooreamatrix)
matrixtest2 <- t(mooreamatrix)

OTUmoorea <- otu_table(matrixtest2, taxa_are_rows = TRUE)
chao1moorea <- estimate_richness(OTUmoorea, measures=c("Chao1"))

#binding the df and removing the incorrect line with 41 (weird mobile bommie that snuck in)
chaomoorea <- cbind(chao1moorea, MooreaH1)
chaomoorea <- chaomoorea[-41,]

```

#### St. John
```{r}
#ST JOHN
stJmatrixt <- t(stJmatrix)
OTUstj <- otu_table(stJmatrixt, taxa_are_rows = TRUE)

chao1stj <- estimate_richness(OTUstj, measures=c("Chao1"))

#binding the df 
chaostj <- cbind(chao1stj, StJohnH1)
```

### percent cover data
#### Moorea
```{r, Moorea percent cover}
MooreaPerCov <- read_csv(here("Data", "MooreaPerCov_H1.csv")) %>%
  mutate(site=as.factor(site))

#make new df that joins the Chao index calculation to the percent cover data
MooreaBenthic <- left_join(MooreaPerCov, chaomoorea, by = c("ID2", "site"))

#mobile bommie is back, need to remove this again
MooreaBenthic <- MooreaBenthic[-41,]
```

#### St. John
```{r, St. John percent cover}
STJPerCov <- read_csv(here("Data", "STJPerCov_H1.csv")) %>%
  mutate(site=as.factor(site))

STJPerCov[is.na(STJPerCov)] = 0

#make new df that joins the Chao index calculation to the percent cover data
STJBenthic <- left_join(STJPerCov, chaostj, by = c("ID", "site", "date"))
```


## Figure 2 - Speces Area Relationship

```{r, figure 2}
Fig2a <- 
  ggplot(data = chaomoorea, aes(x = log(SAm +1), y = log(Chao1+1))) + 
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point()+
  xlim(0, 2.8) +
  ylim(1.6, 4.5) +
  labs(title="a. Moorea", x=expression(paste("log(Surface Area of bommie ", (m^2))), y="log(Chao1)") +
  geom_smooth(method = "lm", color = "#444444")

Fig2b <- 
  ggplot(data = chaostj, aes(x = log(SAm +1), y = log(Chao1+1))) + 
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point()+
  xlim(0, 2.8) +
  ylim(1.6, 4.5) +
  labs(title="b. St. John", x=expression(paste("log(Surface Area of bommie ", (m^2))), y="") +
  geom_smooth(method = "lm", color = "#444444")

Fig2a + Fig2b

```

and calculate standard error for these slopes to report
```{r, SAR standard error}
# Moorea
moorea1 <- lm(log(Chao1+1)~log(SAm +1), data = MooreaBenthic)

coef(moorea1)
moorea1_summary <- summary(moorea1)
moorea1_standard_errors <- moorea1_summary$coefficients[, "Std. Error"]
moorea1_standard_errors

# St. John
stj2 <- lm(log(Chao1+1)~log(SAm +1), data = STJBenthic)

coef(stj2)
stj2_summary <- summary(stj2)
stj2_standard_errors <- stj2_summary$coefficients[, "Std. Error"]

```



## nMDS
### Moorea plot (Appendix S1)
```{r, Moorea nMDS}
# nmds plot showing bommies as points, species as vectors, size and site as colors and polygons

NMDS.log<-log1p(MooreaBenthic[,-1:-22]) #log + 1 of counts
tot <- rowSums(NMDS.log)
NMDS.log <- NMDS.log[tot > 0, ]
sol <- metaMDS(NMDS.log,k=3, distance='bray')
sol$stress

scrs <- as.data.frame(vegan::scores(sol, display = "sites"))
#attach the information of each bommie back to each row 
scrs <- cbind(scrs, Group = c(MooreaBenthic$ID2)) #bommie ID
scrs <- cbind(scrs, Site = c(MooreaBenthic$site)) #site ID
scrs <- cbind(scrs, SAm = c(MooreaBenthic$SAm)) #surface area
scrs <- cbind(scrs, Chao1 = c(MooreaBenthic$Chao1)) #Chao 1 value

#species vectors
vf <- envfit(sol, NMDS.log, perm = 999)

spp.scrs <- as.data.frame(vegan::scores(vf, display = "vectors"))
spp.scrs <- spp.scrs[which(vf$vectors$pvals<= 0.001),]
spp.scrs <- cbind(spp.scrs, Species = rownames(spp.scrs))

scrs$Group <- as.factor(scrs$Group)


#setting numbers to species vectors
genuslabels <- c(row.names(spp.scrs))
genuslabels <- as.data.frame(genuslabels) %>%
  mutate(vector = row_number()) %>%
  rename('ID' = genuslabels)

# create the gt table
genustable <- genuslabels %>%
  gt() %>%
  tab_style(style = cell_text(style = "italic"),
    locations = cells_body(columns = c(`ID`)))

genustable <- genustable %>%
  tab_options(table.font.color = "black")

# convert gt table to HTML format
genustable_html <- as.character(as_raw_html(genustable))

# grob - saving the table as a visual image that can be placed next to the nMDS for easy visuals
gt_filename <- "genustable.png"
gtsave(genustable, filename = gt_filename)

# load the image as a grob
genustable_image <- image_read(gt_filename)
genustable_grob <- rasterGrob(genustable_image, interpolate = TRUE)

# plot
library(viridis)
p <- ggplot(scrs) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, size = SAm, color = Chao1), alpha = 0.7) +
  scale_color_viridis(option = "magma", trans = "reverse") +
  coord_fixed() + ## need aspect ratio of 1!
  geom_segment(data = spp.scrs,
                aes(x = 0 + 1.7, xend = NMDS1 + 1.7, y = 0 + 1.5, yend = NMDS2 + 1.5),
                linewidth = 0.5, color = "darkgrey") +
   geom_text_repel(data = spp.scrs,
                   max.overlaps = Inf,
                   aes(x = NMDS1 + 1.7, y = NMDS2 + 1.5, label = genuslabels$vector),
             size = 6)+
  theme(panel.background = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),  
    axis.ticks = element_blank())+
  labs(title="a. Moorea", size = expression(paste("bommie surface area ", (m^2))), color = "species richness (Chao1)") +
  annotate("text", x=0.75, y=-1.5, label= paste('2D stress = ', round(sol$stress,3)), size = 4)
p

# put the species vector labels to the left
Mfinal_plot <- ggdraw() +
  draw_plot(p, 0, 0, 0.7, 1) +
  draw_plot(genustable_grob, 0.7, 0, 0.3, 1)

Mfinal_plot
```
### St. John plot (Appendix S1)
```{r, St. John nMDS}
# nmds plot showing bommies as points, species as vectors, site as colors and polygons

#check for NAs
csum <- colSums(chaostj[,-1:-15])
# check if any are missing
any(is.na(csum))

stj_NMDSlog<-(log(chaostj[,-1:-15]+1)) #log + 1 of counts (ERROR - log1p wasnt working?)
stj_tot <- rowSums(stj_NMDSlog)
stj_NMDSlog <- stj_NMDSlog[stj_tot > 0, ]
stj_sol <- metaMDS(stj_NMDSlog,k=3, distance='bray')
stj_sol$stress

stj_scrs <- as.data.frame(vegan::scores(stj_sol, display = "sites"))

#attach the information of each bommie back to each row 
stj_scrs <- cbind(stj_scrs, Site = c(chaostj$site))
stj_scrs <- cbind(stj_scrs, SAm = c(chaostj$SAm))
stj_scrs <- cbind(stj_scrs, Chao1 = c(chaostj$Chao1))

#species vectors
stj_vf <- envfit(stj_sol, stj_NMDSlog, perm = 999)

#vectors
stj_spp.scrs <- as.data.frame(vegan::scores(stj_vf, display = "vectors"))
stj_spp.scrs <- stj_spp.scrs[which(stj_vf$vectors$pvals<= 0.001),]
stj_spp.scrs <- cbind(stj_spp.scrs, stj_Species = rownames(stj_spp.scrs))
stj_scrs$Site <- as.factor(stj_scrs$Site)

# setting numbers to species vectors
stj_genuslabels <- c(row.names(stj_spp.scrs))
stj_genuslabels <- as.data.frame(stj_genuslabels) %>%
  mutate(vector = row_number()) %>%
  rename('ID' = stj_genuslabels)

#i want a table in the plot with genera labeled in numbers
#table with black text, species names in italics

# create the gt table
stj_genustable <- stj_genuslabels %>%
  gt() %>%
  tab_style(style = cell_text(style = "italic"),
    locations = cells_body(columns = c(`ID`)))

stj_genustable <- stj_genustable %>%
  tab_options(table.font.color = "black")

# convert gt table to HTML format
stj_genustable_html <- as.character(as_raw_html(stj_genustable))

# grob
stj_gt_filename <- "stj_genustable.png"
gtsave(stj_genustable, filename = stj_gt_filename)

# load image as a grob
stj_genustable_image <- image_read(stj_gt_filename)
stj_genustable_grob <- rasterGrob(stj_genustable_image, interpolate = TRUE)

# plot
stj_p <- ggplot(stj_scrs) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, colour = Chao1, size = SAm), alpha = 0.7) +
  scale_color_viridis(option = "magma", trans = "reverse") +
  coord_fixed() + ## need aspect ratio of 1!
  geom_segment(data = stj_spp.scrs,
               aes(x = 0 + 1.6, xend = NMDS1 + 1.7, y = 0, yend = NMDS2),
               linewidth = 0.5, colour = "darkgrey") +
  geom_text_repel(data = stj_spp.scrs,
                   max.overlaps = Inf,
                   aes(x = NMDS1 + 1.7, y = NMDS2, label = stj_genuslabels$vector),
             size = 6)+
  theme(panel.background = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),  
    axis.ticks = element_blank(),
    legend.position = "bottom")+
  labs(title="b. St. John", size = expression(paste("bommie surface area ", (m^2))), color = "species richness (Chao1)") +
  annotate("text", x=0.2, y=-0.7, label= paste('2D stress = ', round(stj_sol$stress,3)), size = 4)
stj_p

# put the species vector labels to the left
STJfinal_plot <- ggdraw() +
  draw_plot(stj_p, 0, 0, 0.7, 1) +
  draw_plot(stj_genustable_grob, 0.7, 0, 0.3, 1)

STJfinal_plot

```
## GLMs
### Moorea
```{r, Moorea GLMs}
# testing: species richness (predictor variable) ~ surface area (fixed factor), rugosity and percent cover (response variables) 

# rugosity conversion - taking the average of the length and width rugosity measurements of each bommie 
## add column that is average rugosity (Cavg) measured by conformed/planar 
MooreaBenthic$Cavg <- (MooreaBenthic$Rl / MooreaBenthic$l)

# moving lcoaiton of that column in df for better visualization
MooreaBenthic <- relocate(MooreaBenthic, Cavg, .before = NULL, .after = Cwidth) 

# build each model
moorea1 <- lm(log(Chao1+1)~log(SAm +1), data = MooreaBenthic)
moorea2 <- lm(log(Chao1+1)~log(SAm +1)+log(Cavg+1), data = MooreaBenthic)
moorea3 <- lm(log(Chao1+1)~log(SAm +1)*log(Cavg+1), data = MooreaBenthic)
moorea4 <- lm(log(Chao1+1)~log(SAm +1)+log(bare_CCA+1), data = MooreaBenthic)
moorea5 <- lm(log(Chao1+1)~log(SAm +1)*log(bare_CCA+1), data = MooreaBenthic)
moorea6 <- lm(log(Chao1+1)~log(SAm +1)+log(bare_CCA+1)+log(Cavg+1), data = MooreaBenthic)
moorea7 <- lm(log(Chao1+1)~log(SAm +1)*log(bare_CCA+1)*log(Cavg+1), data = MooreaBenthic)

# model selection
AIC(moorea1, moorea2, moorea3, moorea4, moorea5, moorea6, moorea7)
#moorea5 is best fit

#check assumptions
plot(moorea5)

# Pearson's correlations
cor.test(MooreaBenthic$Cavg, MooreaBenthic$SAm)
cor.test(MooreaBenthic$Cavg, MooreaBenthic$bare_CCA)
cor.test(MooreaBenthic$SAm, MooreaBenthic$bare_CCA)

# model output
summary(moorea5)
```

### St. John
```{r, St. John GLMS}

stj2 <- lm(log(Chao1+1)~log(SAm +1), data = STJBenthic)
stj3 <- lm(log(Chao1+1)~log(SAm +1)+log(bare_CCA+1), data = STJBenthic)
stj4 <- lm(log(Chao1+1)~log(SAm +1)*log(bare_CCA+1), data = STJBenthic)

AIC(stj2, stj3, stj4)
# stj2 is best fit

# test assumptions
plot(stj2)

# Pearson's correlations
cor.test(STJBenthic$bare_CCA, STJBenthic$SAm)

# model output
summary(stj2)

```
### effects plots (Appendix S1)
```{r}

moorea7_effects <- plot_model(moorea7, sort.est = TRUE)
moorea7_effects

moorea7_effects_custom <- moorea7_effects +
  theme_classic() + # Change theme
  labs(title = "a. Moorea",
       x = "Predictors",
       y = "Estimated Effect") +
  scale_x_discrete(labels = c("SA * C", "SA * AS", "AS", "C", "AS * C", "(SA * AS) * C", "SA")) +
  theme(plot.title = element_text(size = 14),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom") + # Adjust legend position
  scale_color_manual(values = c("gray42", "gray42"))


# St. John
stj4_effects <- plot_model(stj4, sort.est = TRUE)
stj4_effects

stj4_effects_custom <- stj4_effects +
  theme_classic() + # Change theme
  labs(title = "a. St. John",
       x = "",
       y = "") +
  scale_x_discrete(labels = c("AS", "SA * AS", "SA")) +
  theme(plot.title = element_text(size = 14),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom") + # Adjust legend position
  scale_color_manual(values = c("grey42", "grey42"))

moorea7_effects_custom + stj4_effects_custom
```

# Proximity
## Calculating Bray-Curtis
### Moorea
```{r, Moorea Bray-Curtis calculations}
MbraycurtisproxLOG <- vegdist(MproxmatrixLOG)
mooreaBCprox <- as.matrix(MbraycurtisproxLOG)
as.dist(as.matrix(MbraycurtisproxLOG))

mooreaBCproxdf <- as.data.frame(mooreaBCprox)

# match bommie IDs to be the column names again so you know which is which
colnames(mooreaBCproxdf) <- MooreaH1$ID2

mooreaID <- as.data.frame(MooreaH1$ID2)
mooreaBC <- cbind(mooreaID, mooreaBCproxdf)
names(mooreaBC)[1] <- "bommie.ID"

# you can now pull Bray-Curtis values from this df
```

### St. John
```{r, St. John Bray-Curtis calculations}

STJbraycurtisproxLOG <- vegdist(STJproxmatrixLOG)
stjBCprox <- as.matrix(STJbraycurtisproxLOG)
as.dist(as.matrix(STJbraycurtisproxLOG))

stjBCproxdf <- as.data.frame(stjBCprox)

# match bommie IDs to column names again so you know which is which
colnames(stjBCproxdf) <- StJohnH1$ID

stjID <- as.data.frame(StJohnH1$ID)
stjBC <- cbind(stjID, stjBCproxdf)
names(stjBC)[1] <- "bommie.ID"

# you can now pull Bray-Curtis values from this df
```

## read in and clean data
```{r, nearest neighbor B-C dfs}
BCNNmoorea <- read_csv(here("Data", "BCNNmoorea.csv")) %>%
  mutate(bommie=as.factor(bommie)) %>%
  mutate(NN=as.factor(NN)) 

  
BCNNstjohn <- read_csv(here("Data", "BCNNstjohn.csv")) %>%
  mutate(bommie=as.factor(bommie)) %>%
  mutate(NN=as.factor(NN)) 
```
## GAMs
### Moorea
```{r, Moorea GAM}
mooreaGAM <- gam(BC ~ s(distance_cm, bs = "cr") + 
                   s(SA_difference, bs = "cr"), 
                  data = BCNNmoorea,
                  method = "REML")
mooreaGAMsum <- summary(mooreaGAM)
mooreaGAMsum$s.table[, "edf"] #estimated degrees of freedom
mooreaGAM$df.residual #residual df for model

summary(mooreaGAM)
```

### St. John
```{r, St. John GAM}
stJGAM <- gam(BC ~ s(distance_cm, bs = "cr", k = 9) + 
                s(SA_difference, bs = "cr", k = 9), 
                  data = BCNNstjohn,
                  method = "REML")
stjSAMsum <- summary(stJGAM)
stjSAMsum$s.table[, "edf"] #estimated degrees of freedom
stJGAM$df.residual #residual df for model

summary(stJGAM)
```

#### GAM plots (Appendix S1)
```{r, GAM plots}
p_objmoorea <- plot(mooreaGAM, residuals = TRUE)
p_objmoorea <- p_objmoorea[[1]] #need to take first IV
sm_dfmoorea <- as.data.frame(p_objmoorea[c("x", "se", "fit")])
data_dfmoorea <- as.data.frame(p_objmoorea[c("raw", "p.resid")])

#bind these into one df
fullmooreadf <- cross_join(sm_dfmoorea, data_dfmoorea)

GAMplotMOOREA <- ggplot(fullmooreadf, aes(x = x, y = fit)) +
  geom_point(mapping = aes(x = raw, y = p.resid)) +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se, y = NULL),
              alpha = 0.3) +
  xlim(0, 250) +
  ylim(-0.25, 0.4) +
  geom_line() +
  labs(title="a. Moorea", x="distance between pairs (cm)", y="Partial effect of Bray-Curtis dissimilarity")

# St. John
p_objstj <- plot(stJGAM, residuals = TRUE)
p_objstj <- p_objstj[[1]]
sm_dfstj <- as.data.frame(p_objstj[c("x", "se", "fit")])
data_dfstj <- as.data.frame(p_objstj[c("raw", "p.resid")])

fullstjdf <- cross_join(sm_dfstj, data_dfstj)

GAMplotSTJ <- ggplot(fullstjdf, aes(x = x, y = fit)) +
  geom_point(mapping = aes(x = raw, y = p.resid)) +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se, y = NULL),
              alpha = 0.3) +
  xlim(0, 250) +
  ylim(-0.25, 0.4) +
  geom_line() +
  labs(title="b. St. John", x="", y="")

GAMplotMOOREA + GAMplotSTJ

```

# H2 - Coral diversity is independent of between-bommie habitat patchiness in Moorea
## read and clean data
```{r, centroid data Moorea}
M_fragmentation <-  read_csv(here::here("Data","Moorea_H2.csv"), show_col_types = FALSE) %>%
  mutate(site=as.factor(site)) 

#want average by centroid - want to reduce df first so we only have what we use
centroidGenus <- M_fragmentation[-c(5:10)]

# each genus summarized by centroid
centroidSUM <- centroidGenus %>%
  # Calculate mean by genus in centroid areas
  group_by(centroid, site) %>%
  summarize(fragratio = mean(SA_ratio),
            Acropora=sum(Acropora), 
            Astreopora=sum(Astreopora), 
            Montipora=sum(Montipora), 
            Leptoseris=sum(Leptoseris), 
            Pavona=sum(Pavona), 
            Danafungia=sum(Danafungia), 
            Herpolitha=sum(Herpolitha),
            Lithophyllon=sum(Lithophyllon), 
            Astrea=sum(Astrea),
            Cyphastrea=sum(Cyphastrea), 
            Leptastrea=sum(Leptastrea), 
            Acanthastrea=sum(Acanthastrea), 
            Lobophyllia=sum(Lobophyllia), 
            Pocillopora=sum(Pocillopora), 
            Porites=sum(Porites), 
            Psammocora=sum(Psammocora), 
            Plesiastrea=sum(Plesiastrea),
            Millepora=sum(Millepora)) %>%
  as.data.frame()

availableHabitat <- read_csv(here::here("Data", "Moorea_H2_availableHabitat.csv"), show_col_types = FALSE)

MallFrag <- right_join(availableHabitat, centroiddfShannon, by = c("centroid", "site"))
```

### calculate percent composition by genera
```{r, genera composition}
McentroidVERT <- melt(centroiddfShannon, id=c("centroid", "site", "fragratio", "ShannonMOOREAfrag" ), variable.name = "genera") %>%
  mutate(fragratio = round(fragratio, digits = 1)) %>%
  mutate(fragratio = as.factor(fragratio))

Mpercent_composition_df <- McentroidVERT %>%
  group_by(fragratio) %>%
  mutate(total_count = sum(value)) %>%
  mutate(percent_composition = (value / total_count) * 100) %>%
  select(fragratio, genera, percent_composition, total_count)

Mtotalcount <- Mpercent_composition_df %>%
  group_by(fragratio) %>%
  summarize(total_count = sum(total_count))
```



## GLM
```{r}
# first create a matrix to calculate Shannon diversity
MOOREAfragmatrix <- log(centroidSUM[,-1:-3]+1)

# calculate shannon diversity for each circular survey 
ShannonMOOREAfrag<-diversity(MOOREAfragmatrix, "shannon")

# merge Shannon values to the centroid df
centroiddfShannon <- cbind(centroidSUM, ShannonMOOREAfrag)

# relocate columns for ease of viewing
centroiddfShannon <- relocate(centroiddfShannon, ShannonMOOREAfrag, .before = NULL, .after = fragratio)

#check assumptions
# normality
qqp(centroiddfShannon$ShannonMOOREAfrag, "norm")

# GLM
mooreaFRAG1 <- lm(ShannonMOOREAfrag~fragratio, data = centroiddfShannon)

# homodasticity
bptest(mooreaFRAG1)

summary(mooreaFRAG1)

#view(centroiddfShannon)

```

## Kendall's Tau rank coorealtion
```{r, Kendall's Tau}
cor.test(MallFrag$fragratio, MallFrag$percent_available, method=c("kendall"))
```
## Figure 3a
```{r, Fig 3a}
Fig3a <- 
  ggplot(data = centroiddfShannon, aes(x = fragratio, y = ShannonMOOREAfrag)) + 
  stat_poly_line(geom = "smooth", 
                 method = "lm",
                 se = TRUE,
                 color = "#444444") +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  geom_point()+
  labs(title="a. Moorea", x="fragmentation ratio", y="Shannon-Weiner diversity of corals")

Fig3a
```

## Figure 3c
```{r}
MgeneraPAL2 <- c("#882255", "#BB4781", "#DC8AB3", "#F1BBD6", "#FBD6E8", "#D6CCD1", "#E8BBBB", "#D88888", "#D25252", "#A73C3C", "#751A1A", "#4A0707", "#6F0058", "#AB4B97", "#9873B9", "#FFBBF1", "#DEB5D5", "#D0C9CF")

# plot
Fig3c <- 
ggplot(Mpercent_composition_df, aes(x = fragratio, y = percent_composition, fill = factor(genera))) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = MgeneraPAL2) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12),
        legend.text = element_text(face = "italic")) +
  scale_y_continuous(limits = c(0,101), expand = expansion(mult = c(0.01, 0))) +
  labs(title = "c.",
       x = "fragmentation ratio",
       y = "proportion composition",
       fill = "genera")
Fig3c

```



# H3 - Within-bommie patchiness affects coral diversity in St. John
## read and clean data
```{r}
StJ_fragmentation <- read_csv(here::here("Data", "StJohn_H3.csv"), show_col_types = FALSE) %>%
  mutate(site=as.factor(site)) %>%
  mutate(ID=as.factor(ID))

# percent composition by genera
STJcentroidVERT <- melt(communitydfShannon, id=c("ID", "date", "site", "ShannonSTJfrag", "percent_coral", "perimeter_ratio", "fragratio", "SA_c", "bommie_SA" ), variable.name = "genera") %>%
  mutate(fragratio = round(fragratio, digits = 1)) %>%
  mutate(fragratio = as.factor(fragratio)) %>%
  mutate(SA_c = round(SA_c, digits = 2)) %>%
  mutate(SA_c = as.factor(SA_c))

STJpercent_composition_df <- STJcentroidVERT %>%
  group_by(SA_c) %>%
  mutate(total_count = sum(value)) %>%
  mutate(percent_composition = (value / total_count) * 100) %>%
  select(SA_c, genera, percent_composition, total_count)

STJtotalcount <- STJpercent_composition_df %>%
  group_by(SA_c) %>%
  summarize(total_count = sum(total_count))
```



## GLM
```{r}
# first create a matrix to calculate Shannon diversity
STJfragmatrix <- subset(StJ_fragmentation, select = c(9:ncol(StJ_fragmentation)))

# calculate shannon indices
ShannonSTJfrag<-diversity(STJfragmatrix, "shannon")

# merge Shannon values to the centroid df
communitydfShannon <- cbind(StJ_fragmentation, ShannonSTJfrag)

# relocate columns for ease of viewing
communitydfShannon <- relocate(communitydfShannon, ShannonSTJfrag, .before = NULL, .after = date)


# GLM
STJFRAGmodel <- lm(log1p(ShannonSTJfrag) ~ SA_c, data = communitydfShannon)

# visual assumption check 
plot(STJFRAGmodel)

# output
summary(STJFRAGmodel)
```



## Figure 3b
```{r, Fig 3b}
Fig3b <- 
  ggplot(data = communitydfShannon, aes(x = SA_c, y = ShannonSTJfrag)) + 
  ggpmisc::stat_poly_line(geom = "smooth", 
                 method = "lm",
                 se = TRUE,
                 color = "#444444") +
  ggpmisc::stat_poly_eq(formula = y ~ x, 
               aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~~")), 
              label.x = "right", 
               label.y = "top")+
  geom_point()+
  labs(title="b. St. John", x="fragmentation ratio", y="Shannon-Weiner diversity of corals")

Fig3b

```


## Figure 3d
```{r, Fig 3d}
STJgeneraPAL <- c("#BCC4FF" , "#8A98FF", "#9873B9", "#213AF5", "#0A1DAB", "#000A54", "#00366B", "#2E669C", "#5E9AD4", "#8EBCE8", "#B6CEE4", "#C0C8D0", "#A7CDE0", "#6AAFD2", "#3484AD", "#135F86", "#024263", "#02635F", "#388C88", "#5ABBB6", "#90D4D0", "#44AA99")

Fig3d <- 
ggplot(STJpercent_composition_df, aes(x = SA_c, y = percent_composition, fill = factor(genera))) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = STJgeneraPAL) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12),
        legend.text = element_text(face = "italic")) +
  scale_y_continuous(limits = c(0,101), expand = expansion(mult = c(0.01, 0))) +
  labs(title = "d.",
       x = "fragmentation ratio",
       y = "proportion composition",
       fill = "genera")
```




